A thermodynamic view of dusty protoplanetary disks

### 2017-08-03 appendix.tex
edited in response to AY comments, up to 5.2

Here we consider the more general equation of state for the gas with
$P = K \rhog^{\Gamma}$ where $K$ is a prescribed function of position
and $\Gamma$ is a constant. The effective energy equation is then
Eq. \ref{poly_energy}. % We also include the self-gravitational
% potential $\psi$ of the gas
% plus dust mixture, which satisfies 
% %\begin{align}
% $  \nabla^2 \psi  = 4 \pi G \rho. $
% %\end{align} 
The linearized equations for axisymmetric
disturbances are

\begin{align}
  \ii\sigma\frac{\delta\rho}{\rho} &= \nabla\cdot\dd\bm{v} +
  \dd\bm{v}\cdot\nabla\ln{\rho},\label{lin_mass_full}\\
%  \ii\sigma\frac{\delta P}{P} &= \nabla\cdot\dd\bm{v} +
   \ii\sigma\frac{\delta P}{P} &= \Gamma \nabla\cdot\dd\bm{v} +
  \dd\bm{v}\cdot\nabla\ln{P} - \dd\bm{v}\cdot\nabla\ln{K} - \frac{\dd\mathcal{C}}{P}.\label{lin_energy_full}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \hat{\bm{r}} \cdot \delta\bm{F} -  \hat{\bm{r}} \cdot \nabla\dd\psi ,\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z,\\
  -\ii\sigma\dd v_z &=  \hat{\bm{z}} \cdot \delta\bm{F}  -  \hat{\bm{z}} \cdot \nabla\dd\psi, \\ 
\nabla^2 \delta\psi & = 4\pi G \dd\rho. 
\end{align}  
where the linearizd pressure force $\delta \bm{F}$ is given in
Appendix \ref{lin_press}. These equations do not assume the
radially-local approximation used in numerical computations. From the
linearized meridional momentum equations, we find

\begin{align}
  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV = \int\left( \rho
  \kappa^2 |\dd v_r|^2 + \rho r\frac{\p \Omega^2}{\p z} \dd v_z \dd
  v_r^*  + \ii\sigma \rho \dd \bm{F}\cdot\dd\bm{v}^*
  - \ii \sigma \rho \nabla\dd\psi\cdot  \dd\bm{v}^* 
  \right)dV,\label{meridional}
\end{align}
where the integral is taken over the volume of the fluid. Integrating
by parts and ignoring surface integrals, the last term is
\begin{align}
  \int \ii\sigma\rho \dd \bm{F}\cdot\dd\bm{v}^*dV &=\int\left( \ii\sigma \frac{\dd
    \rho}{\rho}\dd\bm{v}^*\cdot \nabla P - \ii\sigma \dd
  \bm{v}^*\cdot\nabla \dd P  \right)dV\notag\\
&= \int\left( \ii\sigma \frac{\dd
    \rho}{\rho}\dd\bm{v}^*\cdot\nabla P + \ii\sigma \dd
 P  \nabla\cdot \dd \bm{v}^*  \right)dV\notag\\
 & = \int\left[
%   \left(
%   \ii\sigma \frac{\dd P}{P} - \dd\bm{v}\cdot\nabla s + \dd
%   \bm{v} \cdot \nabla \ln{c_s^2} + \frac{\dd\mathcal{C}}{P}
%   \right)\dd\bm{v}^*\cdot\nabla P  
   \left(
   \ii\sigma \frac{\dd P}{\Gamma P} - \dd\bm{v}\cdot\nabla \seff + \frac{\dd
   \bm{v}}{\Gamma} \cdot \nabla \ln{K} +
   \frac{\dd\mathcal{C}}{\Gamma P} \right)\dd\bm{v}^*\cdot\nabla P
   + \ii\sigma \dd P  \nabla\cdot \dd \bm{v}^*
   \right]dV\notag\\
 & = \int\left[
%   \ii\sigma\frac{\dd P}{P}\nabla\cdot\left(P\dd\bm{v}^*\right) +
%   \left(\dd\bm{v}^*\cdot\nabla P\right) \left(\dd\bm{v} \cdot \nabla
%   \ln{c_s^2} + \frac{\dd\mathcal{C}}{P}   -  \dd\bm{v}\cdot\nabla s\right) 
%   \right]dV\notag\\
   \ii\sigma\frac{\dd P}{\Gamma P}\left(\dd\bm{v}^*\cdot\nabla P +
   \Gamma P\nabla\cdot\dd\bm{v}^*   \right)
   +\left(\dd\bm{v}^*\cdot\nabla P\right) \left( \frac{\dd\bm{v}}{\Gamma} \cdot \nabla
   \ln{K} + \frac{\dd\mathcal{C}}{\Gamma P}   -  \dd\bm{v}\cdot\nabla \seff\right) 
   \right]dV\notag\\
 & = \int\left\{
%   \left[\frac{1}{P}\nabla\cdot\left(P\dd\bm{v}\right) -
%   \dd\bm{v}\cdot\nabla\ln{c_s^2} - \frac{\dd\mathcal{C}}{P}  \right]\nabla\cdot\left(P\dd\bm{v}^*\right)
%   +\left(\dd\bm{v}^*\cdot\nabla P\right)\left(\dd\bm{v} \cdot \nabla
%   \ln{c_s^2} + \frac{\dd\mathcal{C}}{P}   -  \dd\bm{v}\cdot\nabla s \right)
%   \right\}dV \notag\\
 \left[
   \nabla\cdot\dd\bm{v} +
  \frac{\dd\bm{v}}{\Gamma}\cdot\nabla\ln{P} -
  \frac{\dd\bm{v}}{\Gamma}\cdot\nabla\ln{K} -
  \frac{\dd\mathcal{C}}{\Gamma P}. 
  \right]
 \left(\dd \bm{v}^*\cdot\nabla P +
 \Gamma P\nabla\cdot\dd\bm{v}^*   \right)\right. \notag\\
 &\phantom{=\int\left\{\right\}}\left. 
 + \left(\dd\bm{v}^*\cdot\nabla P\right)\left(\frac{\dd\bm{v}}{\Gamma} \cdot \nabla
 \ln{K}
 + \frac{\dd\mathcal{C}}{\Gamma P}   -  \dd\bm{v}\cdot\nabla \seff \right)
 \right\}dV \notag\\
 &=\int\left[
%     \frac{1}{P}\left|\nabla\cdot\left(P\dd\bm{v}\right)\right|^2 -
%     \left(\dd\bm{v}^*\cdot\nabla P\right)\left(\dd\bm{v}\cdot\nabla s
%     \right) -
%     P\left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2} + \frac{\dd\mathcal{C}}{P} \right)  
    \frac{1}{\Gamma P} \Bigl\lvert \dd \bm{v}\cdot\nabla P +
 \Gamma P\nabla\cdot\dd \bm{v}    \Bigr\rvert ^2 - 
     \left(\dd\bm{v}^*\cdot\nabla P\right)\left(\dd\bm{v}\cdot\nabla \seff
     \right) -
      P\left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{K} + \frac{\dd\mathcal{C}}{P} \right) 
     \right]dV,
\end{align}
where  $\seff\equiv\ln{(P^{1/\Gamma}/\rho)}$. 
The self-gravitational part of Eq. \ref{meridional} is, again
nelgecting surface terms when integrating by parts, 
\begin{align}
-\int \ii\sigma \rho \nabla\dd\psi \cdot\dd  \bm{v}^* &= \int \ii\sigma
  \dd\psi \nabla\cdot\left(\rho\dd\bm{v}^*\right) dV
 = \int \left|\sigma\right|^2 \dd\psi \dd \rho^*dV 
  = -\frac{1}{4\pi G}\int   \left|\sigma\right|^2
   \left|\nabla\psi\right|^2 dV,  
\end{align}
where the linearized continuity and Poisson equations have been used.

@@

\section{Linearized dust diffusion}\label{lin_dust}
We consider small grains in the Epstein regime, with fixed internal
density and size, so that
\begin{align}
  \tstop  =  \frac{K}{\rho c_s}\label{epstein}
\end{align}
\citep{price15}, where $K$ is a constant. Then the dust diffusion
function becomes
\begin{align}
  \mathcal{C} \equiv c_s^2\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right) = - K c_s^2 \nabla\cdot
  \left(\frac{\tepsilon}{c_s}\bm{F}\right) =
  -Kc_s \left(\bm{F}\cdot\nabla\tepsilon + \tepsilon \nabla\cdot\bm{F}
  - \frac{1}{2}\tepsilon\bm{F}\cdot\nabla\ln{c_s^2}\right).  
\end{align}
Linearizing,
\begin{align}
  - \frac{\dd\mathcal{C}}{Kc_s} = \bm{F}\cdot\nabla\dd\tepsilon +
  \dd\bm{F}\cdot\nabla\tepsilon + \dd\tepsilon\nabla\cdot\bm{F} +
  \tepsilon\nabla\cdot\dd\bm{F} -
  \frac{1}{2}\nabla\ln{c_s^2}\cdot\left(\bm{F}\dd\tepsilon + \tepsilon
  \dd\bm{F}\right). 
\end{align}
The linearized dust-fraction $\dd\tepsilon$ and its derivatives in the
radially-local approximation are given by
\begin{align}
  \dd\tepsilon      &= (1-\tepsilon)W - \frac{Q}{c_s^2}, \\
  \p_r\dd\tepsilon &= \left[(1-\tepsilon)\left(\ii k_x -
    \p_r\ln{\rho}\right) - \p_r\tepsilon\right]W + \left(\p_r\ln{c_s^2}
  + \p_r\ln{\rho} - \ii k_x \right)\frac{Q}{c_s^2},\\
  \dd\tepsilon^\prime &= (1-\tepsilon)W^\prime - \tepsilon^\prime W -
  \left(\frac{Q}{c_s^2}\right)^\prime. 
\end{align}

>>>>

Here we consider the more general equation of state for the gas with
$P = K \rhog^{\Gamma}$ where $K$ is a prescribed function of position
and $\Gamma$ is a constant. The effective energy equation is then
Eq. \ref{poly_energy}, which generalizes
Eq. \ref{eff_energy}. Assuming axisymmetry throughout, we linearize this equation along with Eq. \ref{masseq},  
\ref{momeq},  and \ref{poisson} gives:

% We also include the self-gravitational
% potential $\psi$ of the gas
% plus dust mixture, which satisfies 
% %\begin{align}
% $  \nabla^2 \psi  = 4 \pi G \rho. $
% %\end{align}

\begin{align}
%  \ii\sigma\frac{\delta\rho}{\rho} &= \nabla\cdot\dd\bm{v} +
%  \dd\bm{v}\cdot\nabla\ln{\rho},\label{lin_mass_full}\\
 \ii\sigma\frac{\Delta\rho}{\rho} &= \ii\sigma\frac{\delta\rho}{\rho} - 
  \dd\bm{v}\cdot\nabla\ln{\rho} =  \nabla\cdot\dd\bm{v},\label{lin_mass_full}\\
 %  \ii\sigma\frac{\delta P}{P} &= \Gamma \nabla\cdot\dd\bm{v} +
 % \dd\bm{v}\cdot\nabla\ln{P} - \dd\bm{v}\cdot\nabla\ln{K} - \frac{\dd\mathcal{C}}{P}.\label{lin_energy_full}\\
  \ii\sigma\frac{\Delta P}{P} &= \ii\sigma\frac{\delta P}{P} -
  \dd\bm{v}\cdot\nabla\ln{P} = 
\Gamma \nabla\cdot\dd\bm{v}  - \dd\bm{v}\cdot\nabla\ln{K} - \frac{\dd\mathcal{C}}{P}.\label{lin_energy_full}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \hat{\bm{r}} \cdot \delta\bm{F} -  \hat{\bm{r}} \cdot \nabla\dd\psi ,\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z,\\
  -\ii\sigma\dd v_z &=  \hat{\bm{z}} \cdot \delta\bm{F}  -  \hat{\bm{z}} \cdot \nabla\dd\psi, \\ 
\nabla^2 \delta\psi & = 4\pi G \dd\rho. 
\end{align}  
where the linearizd pressure force $\delta \bm{F}$ is given in
Appendix \ref{lin_press}. Recall $\Delta = \delta +
\bm{\xi}\cdot\nabla$ is the Lagrangian derivative and $\bm{\xi}$ is the
Lagrangian displacement. For axisymmetric flow, we have
$\ii\sigma\bm{\xi}\cdot\nabla = -\dd\bm{v}\cdot\nabla$.  These equations do not assume the
radially-local approximation used in numerical computations. Note that
Eq. \ref{lin_mass_full}---\ref{lin_energy_full} imply 
\begin{align}
\left|\sigma\right|^2\frac{\Delta P}{P}\frac{\Delta \rho^*}{\rho} =
\Gamma\left|\nabla\cdot\dd\bm{v}\right|^2 -
\nabla\cdot\dd\bm{v}^*\left(\dd\bm{v}\cdot\nabla\ln{K} + \frac{\dd\mathcal{C}}{P}\right),
\end{align}
so the first (real) term on the right hand side does not contribute to
$\imag{\left(\Delta P\Delta\rho^*\right)}$. Setting $K$ to constant
and taking the imaginary part gives Eq. \ref{pdv}.

From the
linearized meridional momentum equations, we find  
\begin{align}
  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV = \int\left( \rho
  \kappa^2 |\dd v_r|^2 + \rho r\frac{\p \Omega^2}{\p z} \dd v_z \dd
  v_r^*  + \ii\sigma \rho \dd \bm{F}\cdot\dd\bm{v}^*
  - \ii \sigma \rho \nabla\dd\psi\cdot  \dd\bm{v}^* 
  \right)dV,\label{meridional}
\end{align}
where the integral is taken over the volume of the fluid. Integrating
by parts and ignoring surface integrals, the last term is
\begin{align}
  \int \ii\sigma\rho \dd \bm{F}\cdot\dd\bm{v}^*dV &=\int\left( \ii\sigma \frac{\dd
    \rho}{\rho}\dd\bm{v}^*\cdot \nabla P - \ii\sigma \dd
  \bm{v}^*\cdot\nabla \dd P  \right)dV\notag\\
&= \int\left( \ii\sigma \frac{\dd
    \rho}{\rho}\dd\bm{v}^*\cdot\nabla P + \ii\sigma \dd
 P  \nabla\cdot \dd \bm{v}^*  \right)dV\notag\\
 & = \int\left[
%   \left(
%   \ii\sigma \frac{\dd P}{P} - \dd\bm{v}\cdot\nabla s + \dd
%   \bm{v} \cdot \nabla \ln{c_s^2} + \frac{\dd\mathcal{C}}{P}
%   \right)\dd\bm{v}^*\cdot\nabla P  
   \left(
   \ii\sigma \frac{\dd P}{\Gamma P} - \dd\bm{v}\cdot\nabla \seff + \frac{\dd
   \bm{v}}{\Gamma} \cdot \nabla \ln{K} +
   \frac{\dd\mathcal{C}}{\Gamma P} \right)\dd\bm{v}^*\cdot\nabla P
   + \ii\sigma \dd P  \nabla\cdot \dd \bm{v}^*
   \right]dV\notag\\
 & = \int\left[
%   \ii\sigma\frac{\dd P}{P}\nabla\cdot\left(P\dd\bm{v}^*\right) +
%   \left(\dd\bm{v}^*\cdot\nabla P\right) \left(\dd\bm{v} \cdot \nabla
%   \ln{c_s^2} + \frac{\dd\mathcal{C}}{P}   -  \dd\bm{v}\cdot\nabla s\right) 
%   \right]dV\notag\\
   \ii\sigma\frac{\dd P}{\Gamma P}\left(\dd\bm{v}^*\cdot\nabla P +
   \Gamma P\nabla\cdot\dd\bm{v}^*   \right)
   +\left(\dd\bm{v}^*\cdot\nabla P\right) \left( \frac{\dd\bm{v}}{\Gamma} \cdot \nabla
   \ln{K} + \frac{\dd\mathcal{C}}{\Gamma P}   -  \dd\bm{v}\cdot\nabla \seff\right) 
   \right]dV\notag\\
 & = \int\left\{
%   \left[\frac{1}{P}\nabla\cdot\left(P\dd\bm{v}\right) -
%   \dd\bm{v}\cdot\nabla\ln{c_s^2} - \frac{\dd\mathcal{C}}{P}  \right]\nabla\cdot\left(P\dd\bm{v}^*\right)
%   +\left(\dd\bm{v}^*\cdot\nabla P\right)\left(\dd\bm{v} \cdot \nabla
%   \ln{c_s^2} + \frac{\dd\mathcal{C}}{P}   -  \dd\bm{v}\cdot\nabla s \right)
%   \right\}dV \notag\\
 \left[
   \nabla\cdot\dd\bm{v} +
  \frac{\dd\bm{v}}{\Gamma}\cdot\nabla\ln{P} -
  \frac{\dd\bm{v}}{\Gamma}\cdot\nabla\ln{K} -
  \frac{\dd\mathcal{C}}{\Gamma P}. 
  \right]
 \left(\dd \bm{v}^*\cdot\nabla P +
 \Gamma P\nabla\cdot\dd\bm{v}^*   \right)\right. \notag\\
 &\phantom{=\int\left\{\right\}}\left. 
 + \left(\dd\bm{v}^*\cdot\nabla P\right)\left(\frac{\dd\bm{v}}{\Gamma} \cdot \nabla
 \ln{K}
 + \frac{\dd\mathcal{C}}{\Gamma P}   -  \dd\bm{v}\cdot\nabla \seff \right)
 \right\}dV \notag\\
 &=\int\left[
%     \frac{1}{P}\left|\nabla\cdot\left(P\dd\bm{v}\right)\right|^2 -
%     \left(\dd\bm{v}^*\cdot\nabla P\right)\left(\dd\bm{v}\cdot\nabla s
%     \right) -
%     P\left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2} + \frac{\dd\mathcal{C}}{P} \right)  
    \frac{1}{\Gamma P} \Bigl\lvert \dd \bm{v}\cdot\nabla P +
 \Gamma P\nabla\cdot\dd \bm{v}    \Bigr\rvert ^2 - 
     \left(\dd\bm{v}^*\cdot\nabla P\right)\left(\dd\bm{v}\cdot\nabla \seff
     \right) -
      P\left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{K} + \frac{\dd\mathcal{C}}{P} \right) 
     \right]dV,
\end{align}
where  $\seff\equiv\ln{(P^{1/\Gamma}/\rho)}$. 
The self-gravitational part of Eq. \ref{meridional} is, again
nelgecting surface terms when integrating by parts, 
\begin{align}
-\int \ii\sigma \rho \nabla\dd\psi \cdot\dd  \bm{v}^* &= \int \ii\sigma
  \dd\psi \nabla\cdot\left(\rho\dd\bm{v}^*\right) dV
 = \int \left|\sigma\right|^2 \dd\psi \dd \rho^*dV 
  = -\frac{1}{4\pi G}\int   \left|\sigma\right|^2
   \left|\nabla\psi\right|^2 dV,  
\end{align}
where the linearized continuity and Poisson equations have been used.

@@

\section{Linearized dust diffusion}\label{lin_dust}
We consider small grains in the Epstein regime, with fixed internal
density and size, so that
\begin{align}
  \tstop  \propto  \frac{1}{\rho c_s}\label{epstein}
\end{align}
\citep{price15}. Then the dust diffusion
function becomes
\begin{align}
  \mathcal{C} \equiv c_s^2\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right) = - K c_s^2 \nabla\cdot
  \left(\frac{\tepsilon}{c_s}\bm{F}\right) =
  -Kc_s \left(\bm{F}\cdot\nabla\tepsilon + \tepsilon \nabla\cdot\bm{F}
  - \frac{1}{2}\tepsilon\bm{F}\cdot\nabla\ln{c_s^2}\right).  
\end{align}
Linearizing,
\begin{align}
  - \frac{\dd\mathcal{C}}{Kc_s} = \bm{F}\cdot\nabla\dd\tepsilon +
  \dd\bm{F}\cdot\nabla\tepsilon + \dd\tepsilon\nabla\cdot\bm{F} +
  \tepsilon\nabla\cdot\dd\bm{F} -
  \frac{1}{2}\nabla\ln{c_s^2}\cdot\left(\bm{F}\dd\tepsilon + \tepsilon
  \dd\bm{F}\right). 
\end{align}
The linearized dust-fraction $\dd\tepsilon$ and its derivatives in the
radially-local approximation are given by
\begin{align}
  \dd\tepsilon      &= (1-\tepsilon)W - \frac{Q}{c_s^2}, \\
  \p_r\dd\tepsilon &= \left[(1-\tepsilon)\left(\ii k_x -
    \p_r\ln{\rho}\right) - \p_r\tepsilon\right]W + \left(\p_r\ln{c_s^2}
  + \p_r\ln{\rho} - \ii k_x \right)\frac{Q}{c_s^2},\\
  \dd\tepsilon^\prime &= (1-\tepsilon)W^\prime - \tepsilon^\prime W -
  \left(\frac{Q}{c_s^2}\right)^\prime. 
\end{align}

### 2017-08-03 linear.tex
edited in response to AY comments, up to 5.2

\subsubsection{Orbital frequency} 
From Eq. \ref{steady_momr}} the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

@@

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
 % \epsilon \frac{\mathrm{max}\left(\delta^2,
  %  1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
 \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(1+\epsilon\right)\delta^2},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio dominates 
over that due to the radial temperature gradient for thin dust layers
such that $\delta^2\ll \epsilon$. Otherwise, vertical shear is
associated with the radial temperature gradient. %However, axisymmtric
%\emph{instability} always caused by $\p_rc_s^2$.

@@

\subsection{Instability of dusty edges} 
Here we apply the dusty analog of the Solberg-Hoiland criteria derived
in \S\ref{iso_perfect}. As discussed there, the first criterion in
generally satisifed. Thus we only consider the second condition,
Eq. \ref{dusty_solberg2}.

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, a conclusion valid for arbitrarily thin
dust layers.

@@

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
LHS is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect density gradients in Eq. \ref{spm} 
if $r\p_r\ln{\rhog}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

Technically, the above discussion is only applicable to strictly
isothermal disks perfectly coupled to dust (see \S\ref{iso_perfect}). 
However, since dusty edges translate to sharp entropy gradients (see
\S\ref{dusty_entropy}), we may generally expect sharp features in the
dust distribution of protoplanetary disks to be unstable.

>>>>

\subsubsection{Orbital frequency} 
From Eq. \ref{steady_momr} the disk orbital frequency is 
\begin{align}
  \Omega(r,z) = \OmK(r)\left[1 - \frac{3}{2}\frac{z^2}{r^2} +
    \frac{h_\mathrm{g}^2}{\left(1+\epsilon\right)}\frac{\p}{\p\ln{r}}\ln{\left(c_s^2\rhog\right)}
    \right]^{1/2}, 
\end{align}
where 
\begin{align}
  h_\mathrm{g} \equiv \frac{\Hgas}{r}
\end{align}
is the characteristic disk aspect-ratio.

@@

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
 % \epsilon \frac{\mathrm{max}\left(\delta^2,
  %  1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
 \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(1+\epsilon\right)\delta^2},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio dominates 
over that due to the radial temperature gradient for thin dust layers
such that $\delta^2\ll \epsilon$. %Otherwise, vertical shear is
%associated with the radial temperature gradient. %However, axisymmtric
%\emph{instability} always caused by $\p_rc_s^2$.

@@

\subsection{Instability of dusty edges} 
Here we apply the dusty analog of the Solberg-Hoiland criteria derived
in \S\ref{iso_perfect} assuming strictly isothermal gas. As discussed
there, the first criterion in generally satisifed. Thus we only
consider the second condition, Eq. \ref{dusty_solberg2}.

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

@@

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
LHS is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
% \chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
%   \left[
%   \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
%   \sqrt{
%   \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
%   4 \frac{\Htilde^2}{H_\mathrm{g}^2}
%   \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
%   }
%   \,\right]. 
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{P}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{P}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_rP<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect pressure gradients in Eq. \ref{spm} 
if $r\p_r\ln{P}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio 
is much less its vertical lengthscale, $L_\epsilon\ll O(\Htilde)$, 
then the system is  unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets.

Technically, the above discussion is only applicable when $c_s$ is
constant and $\tstop=0$ (see \S\ref{iso_perfect}). However, since
dusty edges translate to sharp entropy gradients (
\S\ref{dusty_entropy}), we may generally expect sharp features in the
dust distribution of protoplanetary disks to be unstable.
